Hypercoercions and a Framework for Equivalence of Cast Calculi

### 2019-10-24 paper.tex
finish revising sec3

\begin{itemize}
\item threesomes \citep{siek2010threesomes,garcia2013calculating},
\item supercoercions \citep{garcia2013calculating}, and
\item coercions in normal form
  \citep{siek2012interpretations,siek2015blame}.
\end{itemize}
%Recall that in these systems, casts are compressed using a composition
%operator.
These systems compress casts using a compose operator.
Threesomes and supercoercions are good for mechanized
metatheory because their compose operators are structurally recursive,
making them easy to define in a proof assistant such as Agda. In
contrast, the coercions in normal form have compose operators that are
not structurally recursive, which makes it more difficult to define in
proof assistants, requiring what amounts to an explicit proof of termination.
%
On the other hand, coercions in normal form are easier to understand
than threesomes (with a strange labelled bottom type), and
supercoercions (10 different kinds).

@@

This section presents our first contribution, the definition of
hypercoercions.  The design of hypercoercions is motivated by the
observation that a normal coercion has at most three
parts. Hypercoercions make this structure explicit: a hypercion $c$ is
either $\hyperCoercionI$, the identity cast for $\TOOdyn$, or it contains
three parts: a head, middle, and tail, as defined in
\figref{fig:hypercoercion}.
\begin{itemize}
\item The head $h$ is either a projection or the identity,
\item the middle $m$ involves coercions that preserve a type
  constructor, i.e., coercions between function, pair, sum types, and
  unit types, and
\item the tail $t$ is either an injection, a failure, or the identity.
\end{itemize}
There is a close correspondence between middle coercions and type
constructors.  We generalize shallow-consistency to middles $m \smile
m$ in the obvious way.

@@

The $seq$ operator is defined in terms of coercion composition, which
is the key to compressing coercions and maintaining space efficiency.
A composition operator for coercions typically requires the target
type of the first coercion to match the source type of the second.
However, it is useful to relax this restriction for the D blame
tracking strategy.
%
We shall need optional blame labels, written $\ell$ , that range over
$\epsilon$ or $l$. The label is mandatory when the target
type of the first coercion does not match the source type of the second. 
Then we write composition of hypercoercions as $c_1
\fatsemi^\ell c_2$.  When both $c_1$ and $c_2$ are $\hyperCoercionI$,
their composition is also $\hyperCoercionI$.  If the head of $c_2$ is
a projection, the result is $c_2$.  Otherwise, we need a label to
build the projection. Since the head of $c_2$ is $\epsilon$, its
source type must be a pretype. Thus we know $\ell$ must be a label and
we put it in the projection. When $c_1$ ends with a failure, the
composition is $c_1$ itself. In all the remaining cases, $c_1$ does
not end with a failure.  When $m_1$ and $m_2$ have different top
constructors ($m_1 \not\smile m_2$), the result is a failure coercion,
which needs a label. When $c_2$ starts with a projection, we blame the 
projection for casting a value to a shallowly inconsistent type. When $c_2$
starts with a $\epsilon$, we know $\ell$ must be a label, which is blamed for
composing shallowly inconsistent hypercoercions. The last two cases
compose $m_1$ and $m_2$ with the auxilliary function $m \fatsemi^\ell
m$, which assumes its inputs have the same top constructor. The
definition of $m \fatsemi^\ell m$ is a straightforward structural
recursion. Going back to the last two cases of $c_1 \fatsemi^\ell
c_2$. When $c_2$ starts with a projection, we do not know whether the
target type of $m_1$ matches the source of $m_2$. So we use the $l$ in the 
projection to  compose the middles $m_1 \fatsemi^l m_2$. When $c_2$ starts with 
a $\epsilon$, we compose the middles using $\ell$.

@@

\item If the middle is of base type, then the remaining 62 bits are
  used to represent the head (21 bits), middle (22 bits), and tail (19
  bits).
  \begin{itemize}
  \item The head requires 1 bit to differentiate between $\epsilon$
    and $?^l$ and then 24 bits could be used for the label $l$.
  \item The middle would use 11 bits to differentiate all the base types.
  \item The tail requires 2 bits to differentiate between $\epsilon$,
    $!$, and $\bot^l$, and could use 24 bits for the label $l$.
  \end{itemize}
  In the event that a program requires too many blame labels, then one
  can fallback to using the heap-allocated structure for more
  coercions.
\end{itemize}

>>>>

\begin{itemize}
\item threesomes \citep{siek2010threesomes,garcia2013calculating},
\item supercoercions \citep{garcia2013calculating}, and
\item coercions in normal form
  \citep{siek2012interpretations,siek2015blame}.
\end{itemize}
%Recall that in these systems, casts are compressed using a composition
%operator.
These systems compress casts using a compose operator.
Threesomes and supercoercions are good for mechanized
metatheory because their compose operators are structurally recursive,
making them easy to define in a proof assistant such as Agda. In
contrast, the coercions in normal form have compose operators that are
not structurally recursive, which makes it more difficult to define in
Agda, requiring what amounts to an explicit proof of termination.
%
On the other hand, coercions in normal form are easier to understand
than threesomes (with a strange labelled bottom type), and
supercoercions (10 different kinds).

@@

This section presents our first contribution, the definition of
hypercoercions.  The design of hypercoercions is motivated by the
observation that a normal coercion has at most three
parts. Hypercoercions make this structure explicit: a hypercion $c$ is
either $\hyperCoercionI$, the identity cast for $\TOOdyn$, or it contains
three parts: a head, middle, and tail, as defined in
\figref{fig:hypercoercion}.
\begin{itemize}
\item The head $h$ is either a projection or the no-op,
\item the middle $m$ involves coercions that preserve a type
  constructor, i.e., coercions between function, pair, sum types, and
  unit types, and
\item the tail $t$ is either an injection, a failure, or the no-op.
\end{itemize}
There is a close correspondence between middle coercions and type
constructors.  We generalize shallow-consistency to middles $m \smile
m$ in the obvious way.

@@

The $seq$ operator is defined in terms of coercion composition, which
is the key to compressing coercions and maintaining space efficiency.
A composition operator for coercions typically requires the target
type of the first coercion to match the source type of the second.
However, it is useful to relax this restriction for the D blame
tracking strategy.
%
We shall need optional blame labels, written $\ell$ , that range over
$\epsilon$ or $l$. The label is mandatory when the target
type of the first coercion does not match the source type of the second. 
Then we write composition of hypercoercions as $c_1
\fatsemi^\ell c_2$.  When both $c_1$ and $c_2$ are $\hyperCoercionI$,
their composition is also $\hyperCoercionI$.  If the head of $c_2$ is
a projection, the result is $c_2$.  Otherwise, we need a label to
build the projection. Since the head of $c_2$ is the no-op, its
source type must be a pretype. Thus we know $\ell$ must be a label and
we put it in the projection. When $c_1$ ends with a failure, the
composition is $c_1$ itself. In all the remaining cases, $c_1$ does
not end with a failure.  When $m_1$ and $m_2$ have different top
constructors ($m_1 \not\smile m_2$), the result is a failure coercion,
which needs a label. When $c_2$ starts with a projection, we blame the 
projection for casting a value to a shallowly inconsistent type. When $c_2$
starts with the no-op, we know $\ell$ must be a label, which is blamed for
composing shallowly inconsistent hypercoercions. The last two cases
compose $m_1$ and $m_2$ with the auxilliary function $m \fatsemi^\ell
m$, which assumes its inputs have the same top constructor. The
definition of $m \fatsemi^\ell m$ is a straightforward structural
recursion. Going back to the last two cases of $c_1 \fatsemi^\ell
c_2$. When $c_2$ starts with a projection, we do not know whether the
target type of $m_1$ matches the source of $m_2$. So we use the $l$ in the 
projection to  compose the middles $m_1 \fatsemi^l m_2$. When $c_2$ starts with 
the no-op, we compose the middles using $\ell$.

@@

\item If the middle is of base type, then the remaining 62 bits are
  used to represent the head (25 bits), middle (11 bits), and tail (26
  bits).
  \begin{itemize}
  \item The head requires 1 bit to differentiate between $\epsilon$
    and $?^l$ and then 24 bits could be used for the label $l$.
  \item The middle would use 11 bits to differentiate all the base types.
  \item The tail requires 2 bits to differentiate between $\epsilon$,
    $!$, and $\bot^l$, and could use 24 bits for the label $l$.
  \end{itemize}
  In the event that a program requires too many blame labels, then one
  can fallback to using the heap-allocated structure for more
  coercions.
\end{itemize}
